\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{booktabs}
\hypersetup{colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue}

\title{Cracking ARC-AGI-3 Games with a Self-Improving Agent\\
under a Free-Energy Selection Principle}
\author{GKM Project}
\date{2026}

\newcommand{\F}{\mathcal{F}}

\begin{document}
\maketitle

\begin{abstract}
We study whether an agent can \emph{figure a game out on its own} --- discover its
perception, its mechanics, its goal, and a winning strategy --- from the rawest
possible interface, carrying only general human preconceptions, with a single
free-energy rule deciding what structure is worth keeping. The setting is ARC-AGI-3
keyboard games played offline through their runtime. Our agent is given nothing but
\texttt{step(action)}$\to$\emph{frame}, a scalar reward (\emph{levels completed}), and
a clone for lookahead. A proposer, prompted with human world-priors, \emph{writes its
own program} to play; a candidate is admitted only if it verifiably improves on the
incumbent under a free energy $\F = R + \lambda C$ that trades reward against
description length. We connect this to Kolmogorov complexity and Solomonoff induction
(the $C$ term), to Schmidhuber's G\"odel machine and PowerPlay (self-rewriting under an
empirical proof of improvement, driven by compression-progress novelty), and to a
colimit-cone view of skills (``legs'') composed into solutions. Empirically, on
\texttt{wa30} the agent wrote a solver clearing levels~1--6 (replay-validated),
finding level tactics beyond its priors --- though we identify, and retire, a bias:
those runs' priors named the carry mechanic and hinted the relay, distilled from
earlier human play; the same
game-agnostic agent transferred to a second game, \texttt{ls20}, clearing levels~1--4.
When the harness enforces a growing \emph{leg library}, the marginal description length
of each new level collapses ($55\to18\to2\to2$ on \texttt{ls20}), i.e.\ later levels are
solved by \emph{composing} existing legs with almost no new structure --- transfer,
priced by the same free energy; on \texttt{wa30}, whose levels introduce genuinely new
mechanics, marginal novelty instead tracks the novelty the game demands
($112\to78\to95$), as a novelty-priced free energy should. We report the honest
negatives alongside: which levels remain unsolved, and that a prompt-only local model
is currently too weak to drive the same loop.
\end{abstract}

\section{Introduction}
ARC-AGI-3~\cite{chollet2019measure,arcagi3} poses interactive grid games whose rules are not
given: an agent sees a $64\times64$ grid of colour integers, may press one of a few
opaque buttons, and receives only a coarse reward when it completes a level. Nothing is
labelled --- not the avatar, not the objects, not the goal. Solving such a game from
scratch requires \emph{discovering} the perception, the manipulation mechanics, the
objective, and a plan, and then \emph{transferring} what was learned to later levels
and other games.

Our thesis is that an agent equipped with a rich stock of \emph{human preconceptions}
about the world, and disciplined by a single \emph{free-energy} selection rule, can do
this itself, with the human contribution confined to (i) the rawest game-independent
interface, (ii) the preconception priors, and (iii) the free-energy accounting. We make
three points precise. First, the interface must expose \emph{only} what is common to
all games, or it smuggles in a solution that will not transfer. Second, the reusable
structure that lets later levels be cheap must be \emph{invented by a proposer}, not
enumerated from a fixed library --- and admitted only against an empirical proof.
Third, the accounting that decides ``is this structure worth its cost'' is the same
free energy throughout, and it is exactly the object that makes transfer the parsimonious
choice.

\section{Background}
\subsection{Kolmogorov complexity, MDL, and free energy}
The (prefix) Kolmogorov complexity $K(x)$ is the length of the shortest program that
outputs $x$; Solomonoff induction weights hypotheses by $2^{-K}$, formalising Occam's
razor~\cite{kolmogorov1965,solomonoff1964,live2008}. $K$ is uncomputable, but its
description-length spirit is usable: prefer the shortest structure that explains the
data. We adopt a loss--complexity \emph{free energy}
\begin{equation}
  \F_\lambda(s) \;=\; R(s) \;+\; \lambda\,C(s),
  \label{eq:fe}
\end{equation}
where $R$ is empirical goal risk (here, unsolved levels) and $C$ is a description-length
complexity of the candidate structure $s$; $\lambda$ tunes the pressure toward compact
solutions~\cite{losscomplexity}. Equation~\eqref{eq:fe} is an MDL objective: a
structure earns its complexity only if it buys enough reduction in risk.

\subsection{Schmidhuber: self-rewriting, PowerPlay, curiosity}
Schmidhuber's G\"odel machine~\cite{schmidhuber2007godel} is an agent that may rewrite
\emph{any} part of its own code, including its learning algorithm, once it finds a proof
that the rewrite increases expected reward. PowerPlay~\cite{schmidhuber2013powerplay}
incrementally searches for the simplest task the system cannot yet solve together with a
solver for it, never forgetting old skills. Artificial curiosity rewards
\emph{compression progress}: transitions that make the agent's world model more
predictable~\cite{schmidhuber2010curiosity}. We take three ideas: (a) the agent should
\emph{write and revise its own problem-solving code}; (b) a candidate is adopted only
against a \emph{proof of improvement} --- which we discharge \emph{empirically}, by the
simulator, rather than by a theorem prover; (c) novelty/compression-progress can
\emph{steer} exploration but is priced, not free.

\subsection{The colimit-cone view of skills}
A recurring construction in this project treats few-shot competence as a diagram of
reusable behaviours (``legs'') together with priced gluings (``cones'') that compose
them into a solution, selected by free energy. In the classical form the legs are
\emph{discovered by interaction} and the cone is a search over them. Here we push the
construction to its limit: the legs are \emph{written by the proposer} as program
fragments, the cone that glues them is the proposer's own planner, and admission remains
Eq.~\eqref{eq:fe}. The essential invariant is that a reused leg is already paid for, so
composition is cheap and only genuine novelty is charged.

\section{Method}
\subsection{The rawest substrate}
A persistent failure mode is smuggling the solution into the harness as a ``primitive'':
any helper richer than the bare runtime --- a perception module, a typed object model, a
push-versus-carry mechanic, a search planner --- is the programmer solving the game, and
it does not survive a change of game \emph{type}. We therefore expose only
\begin{center}
\texttt{step(action)}$\to$\emph{frame}, \quad \emph{frame} (raw $64\times64$ grid),
\quad \emph{levels\_completed} (reward), \quad \texttt{clone()} (lookahead).
\end{center}
Nothing here names objects, colours, walls, avatars, or goals. Perceiving objects,
finding the controllable object, learning the mechanic, inducing the goal, modelling other
agents, planning, and composing manoeuvres must all be produced \emph{inside} the loop.

\subsection{Preconceptions and the proposer}
The proposer receives a system prompt of \emph{human preconceptions}: that frames contain
persistent objects; that barriers partition space into regions and reachability must be
reasoned per agent; that one object may be a controllable avatar and others autonomous
agents (helpful or adversarial) that act on every move; that cooperation can require a
hand-off at a shared boundary; that rewards are often sparse, so the agent should set
itself dense surrogate objectives and discover its actions' affordances by experiment.
These priors are deliberately \emph{non-procedural}: they say what a world tends to
contain, not how to solve any game. Given the priors and the raw substrate the proposer
\emph{writes a} \texttt{solve(env)} \emph{program} --- its perception, its mechanic
probe, its planner, its strategy. The proposer is pluggable: a small local model, or the
Claude Code agent invoked headlessly with file/shell tools and a tester, so it writes,
runs, observes failures, and iterates --- the same write/run/fix loop a programmer uses.
All reported results use the strong proposer; the small local model is presently too
weak to drive the loop (Section~\ref{sec:discussion}).

\subsection{Admission by free energy; the leg library}
A written program is admitted only if it verifiably lowers Eq.~\eqref{eq:fe} on the real
game: $R=-(\text{levels reached})$ and $C$ its description length, with $\lambda$ small so
parsimony breaks ties without stifling a larger program that clears more levels;
compression-progress/disagreement add an intrinsic novelty term that only \emph{steers}
exploration; earlier behaviour is preserved against regression; and the simulator is the
ground-truth verifier (a candidate is replayed on a fresh environment and kept only if
the reward actually advances). Two constraints govern the loop: selection cannot be a
\emph{source} of novelty (that is the proposer's job), and discovery/strategy must be
learned inside the loop, never hand-coded.

To make later levels cheap we grow a \emph{leg library}. A shared module accumulates
small named skills (perception helpers, navigation, transport, relay-across-a-barrier);
each level's player is restricted to \emph{composing} those legs; after each level a
\emph{debrief} refactors repeated code into shared legs and logs the recurring
composition (itself a candidate higher-order leg). Crucially, the complexity term is
\emph{marginal}: $C_{\text{marg}}$ counts only \emph{new} legs plus composition glue, so a
reused leg costs nothing and Eq.~\eqref{eq:fe} literally rewards transfer. This is the
colimit cone made operational, and it must be \emph{enforced by the harness}: merely
asking the proposer to grow a library is ignored (it grows a monolith); restricting each
player to composition over a shared library, with a separate refactor pass and marginal
scoring, produces the intended behaviour.

\section{Results}
All results are replay-validated: the recorded action sequence is re-run on a fresh
environment and must reach the claimed level.

\paragraph{Discovery and depth (\texttt{wa30}).} Handed only the raw substrate and the
priors, the strong proposer wrote a single adaptive \texttt{solve(env)} that cleared
\texttt{wa30} levels~1--3, and a continued run reached level~6 (a $458$-move sequence,
$\F=-3.903$). Levels~7--9 remain unsolved. An honest audit of \emph{what was given}
is owed here. The priors used in these runs were not fully neutral: distilled from
earlier human play of this same game, they contained an attach$\to$carry$\to$release
experiment recipe and a relay-at-a-shared-boundary hint, and the interaction probe
handed the proposer a mechanic \emph{name} (``pick-up-and-carry'') from a hand-coded
verb vocabulary. So the mechanic family and the relay \emph{concept} were pre-told.
What the agent demonstrably found on its own, beyond the priors: it froze the target
region at level start (delivered objects take on the container colour and vanish from
a naive detector); it \emph{complemented} the autonomous helper by taking the objects
\emph{farthest} from the goal rather than competing; and it discovered the
\emph{asymmetric} carry collision --- a carried object can enter a wall cell the
avatar cannot --- that makes the relay geometrically possible at all. We have since
\emph{neutralized} the priors (generic world-priors only: no mechanic recipes, no
verb names, no relay hint); re-cracking \texttt{wa30} from scratch under the neutral
priors is the discriminating experiment.

\paragraph{Transfer across game type (\texttt{ls20}).} The \emph{same} game-agnostic
agent, pointed at \texttt{ls20} --- a slide-to-match game with a different mechanic --- and
with \emph{no code change}, discovered \texttt{ls20}'s own perception, mechanic, and goal
and cleared levels~1--4. This is the transfer the rawest substrate is designed for.

\paragraph{Leg reuse and marginal novelty.} Running \texttt{ls20} under the enforced leg
library (capped at level~4) yields the signature predicted by Eq.~\eqref{eq:fe} with a
marginal complexity term (Table~\ref{tab:marg}). Level~1 invents the library; levels~3
and~4 add essentially nothing and are one call to the level-advancing leg; level~2 reuses
the same leg with a single knob (a fuller state key, because its goal depends on carried
objects), threaded in by the debrief so both players share one generic search. Total
marginal novelty for four levels is $77$, almost all front-loaded --- transfer, not
re-derivation, and priced as such.

\begin{table}[h]
\centering
\begin{tabular}{lccc}
\toprule
level & marginal $C$ (new structure) & cumulative $C$ & $\F$ \\
\midrule
1 & 55 \; (invents the legs) & 55 & $+0.10$ \\
2 & 18 \; (reuse $+$ one knob) & 73 & $-0.54$ \\
3 & \phantom{0}2 \; (pure composition) & 75 & $-1.50$ \\
4 & \phantom{0}2 \; (pure composition) & 77 & $-2.46$ \\
\bottomrule
\end{tabular}
\caption{Enforced leg library on \texttt{ls20}: marginal description length per level
collapses as legs are reused ($\lambda=0.02$, $R=-\text{level}$).}
\label{tab:marg}
\end{table}

\paragraph{The complement on \texttt{wa30}: novelty tracks the game's novelty.} Running
\texttt{wa30} under the same enforced library (levels 1--3, replay-validated) shows the
honest counterpart: marginal novelty does \emph{not} collapse ($112 \to 78 \to 95$),
because each \texttt{wa30} level introduces a genuinely new mechanic --- level~2 an
autonomous helper agent, level~3 a dividing wall crossed only by an asymmetric carry
collision --- and the agent captured each as new named legs (\texttt{yield\_to\_helper},
\texttt{wall\_col}, \texttt{relay\_to\_helper}). Reuse-collapse is therefore a property
of the \emph{game's} level structure, not an artifact of the harness: the marginal price
in Eq.~\eqref{eq:fe} is paid exactly when the game demands new structure and drops to
composition cost when it does not, which is the intended behaviour of a novelty-priced
free energy. The library discipline still holds --- players remain thin composers
(25 lines over a 24-leg library), and the level-3 wall relay was discovered \emph{inside}
the enforced loop and persists as a reusable skill.

\section{Discussion and honest limitations}\label{sec:discussion}
The architecture --- rawest substrate, preconception priors, free-energy admission,
simulator-as-verifier, marginally-priced leg library --- is general and is the
contribution. Two honest bounds apply. (1) The claim is levels~1--6 of \texttt{wa30}
and~1--4 of \texttt{ls20}, not all levels and not all games. (2) The loop currently needs
a strong proposer: a prompt-only local model mis-reasoned two-sided reachability under
barriers even with the priors spelled out, so the interesting open question is how weak a
proposer the same harness --- priors, simulator-as-verifier, free-energy admission ---
can lift to competence.
Relative to the classical colimit-cone experiments, where reusable structure is enumerated
and priced, here it is \emph{invented} and priced by the same free energy; the two are
ends of one spectrum, and Table~\ref{tab:marg} is direct evidence that on the invented
end, transfer is the parsimonious choice.

\bibliographystyle{plain}
\bibliography{references}
\end{document}
