\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{hyperref}

\title{Bongard Semantic Cones}
\author{Alexander Kolpakov}
\date{}

\begin{document}
\maketitle

\begin{abstract}
The Bongard system in this repository has been extended with typed semantic
cones alongside the existing unrestricted predicate-search control.  The target
of the semantic track is not merely to find reusable differentiators over pixels.
The target is to recover human-like semantic descriptions whose executable
decisions factor through declared objects, bindings, relations, and nuisance
morphisms.  Kolmogorov/MDL selection remains the outer selector, but only among
admissible semantic cones.
\end{abstract}

\section{Problem}

A Bongard task contains a small positive class and a small negative class.
Many functions can separate twelve panels.  Some are genuine semantic
descriptions; many are accidental differentiators.  A reusable differentiator
may be Kolmogorov-cheap while still failing to match the human concept.  The
system must therefore distinguish semantic recovery from unrestricted
classification.

\section{Semantic Cone}

Let $\mathcal P$ be an operational category of panel representations and
transformations.  Let $\mathcal S$ be a typed relational category containing
scenes, objects, descriptors, measurements, relations, and Booleans.  A
semantic realization is a functor-like map
\[
S : \mathcal P \to \mathcal S .
\]

A hypothesis is a typed diagram $D_H : J \to \mathcal S$.  A solution is an
executable cone over that diagram.  The final predicate must factor as
\[
p_H = \beta \circ r \circ b \circ S,
\]
where $b$ selects and binds variables, $r$ evaluates declared measurements or
relations, and $\beta$ evaluates the Boolean rule.  A direct panel function
$q : \mathrm{Panel}\to\mathrm{Bool}$ is not a semantic solution unless this
factorization is exhibited by the compiler.

\section{Human-Like Constraint}

The artifact records the selected description.  A cone that uses only a direct
panel statistic, such as total ink, is reported as a measurement-only control,
not as a semantic-pure solve.  Semantic-pure admission requires an object or
relation factorization and declared preservation morphisms.  This does not
prove that the English sentence is uniquely human, but it rejects the known
failure in which natural language and code are merely paired rather than
causally connected.

\section{Verification}

Thresholds are fit by leave-one-out folds.  Declared nuisance morphisms define
naturality obligations, and contrast interventions define discriminative
obligations.  The verifier should report the first noncommuting node when a
cone fails under transformation.

\section{Selection}

For a cone $C$ and promoted library $L$, the objective is
\[
F(C\mid L) =
R_{\mathrm{support}} + R_{\mathrm{loo}} + R_{\mathrm{naturality}}
+ R_{\mathrm{contrast}} + \lambda K(C\mid L).
\]
Previously admitted legs are not recharged, but calls, bindings, diagrams, and
parameters are charged.  The MDL term selects among admissible semantic cones;
it does not license bypassing the semantic factorization.

\section{Implementation Status}

The active code paths are in \texttt{bongard/crack\_lab}.  The unrestricted
\texttt{predicates.py} crack loop remains the control track.  The semantic
runner accepts cofibered proposer outputs: typed cones over the current problem
object.  Static cones are allowed only as unit test fixtures.  Hybrid solutions,
when added, must label residual guards separately and price them explicitly.

\bibliographystyle{plain}
\bibliography{references}

\end{document}
